---
title: 'STAT 420: Final Project'
author: "Chenhan Xu (chenhan2)"
date: 'Friday, May 3'
output:
  html_document:
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
---

**Abstract**

In this report, we will try to fit a proper linear regression model to predict the `Disease.Severity` in the `disease` dataset using other variables. We will first check the multicollinearity of the dataset and try to reduce the predictors. Then we will fit the additive and interaction full linear models and try to interpret and diagnose them with different methods. As these two model do not fit the data well, we try to use response transformation to find a better one. For each family of models, we use Bayesian Information Criterion (BIC) to do the variable selection. We prefer BIC here instead of AIC because with large sample data size, BIC tends to find a smaller model that can be more interpretable. For each model we find using BIC, we will use Cook's Distance to remove the influencial unusual observation. Test of model assumption will be given to each of the models. Finally, this report will arrive at a proper model for the dataset.

**Setup of the project**

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```

Import the `disease` dataset and the packages that contain the necessary functions for this project. 

```{r, warning = FALSE, message = FALSE}
disease = read.csv("Disease Data.csv")
library(lmtest)
library(faraway)
library(MASS)
library(leaps)
```

Define some of the functions that we may need for convenience.

```{r}
fitted_vs_resid = function(model) {
  plot(fitted(model), resid(model),
       col = "grey", pch = 20,
       xlab = "Fitted Value",
       ylab = "Residual",
       main = "Fitted Value versus Residual")
  abline(h = 0, lwd = 3, col = "dodgerblue")
}
qqplot = function(model) {
  qqnorm(resid(model), col = "grey", lwd = 3)
  qqline(resid(model), col = "dodgerblue", cex = 1.5, pch = 20)
}
calc_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}
```


```{r}
summary(disease)
```

```{r}
str(disease)
```



```{r}
disease = subset(disease, select = - X)
```

```{r}
#pairs(disease)
correlation_within = cor(disease)
```

```{r}
abs(correlation_within) > 0.8
```

```{r}
vif(disease)
vif(disease)[which(vif(disease) > 5)]
```
```{r}
fit_full_without_F2R = lm(Disease.Severity ~ . - F2R, data = disease)
fit_F2R = lm(F2R ~ . - Disease.Severity, data = disease)
```
```{r}
cor(resid(fit_full_without_F2R), resid(fit_F2R))
```
```{r}
plot(resid(fit_full_without_F2R) ~ resid(fit_F2R),
     col = "dodgerblue", pch = 20,
     xlab = "Residuals, Added Predictor", 
     ylab = "Residuals, Original Model")
abline(h = 0, lty = 2)
abline(v = 0, lty = 2)
abline(lm(resid(fit_full_without_F2R) ~ resid(fit_F2R)),
       col = "darkorange", lwd = 2)
```
```{r}
fit_full_without_PLEKHM1 = lm(Disease.Severity ~ . - PLEKHM1, data = disease)
fit_PLEKHM1 = lm(PLEKHM1 ~ . - Disease.Severity, data = disease)
```
```{r}
cor(resid(fit_full_without_PLEKHM1), resid(fit_PLEKHM1))
```
```{r}
plot(resid(fit_full_without_PLEKHM1) ~ resid(fit_PLEKHM1),
     col = "dodgerblue", pch = 20,
     xlab = "Residuals, Added Predictor", 
     ylab = "Residuals, Original Model")
abline(h = 0, lty = 2)
abline(v = 0, lty = 2)
abline(lm(resid(fit_full_without_PLEKHM1) ~ resid(fit_PLEKHM1)),
       col = "darkorange", lwd = 2)
```

```{r}
fit_full_add = lm(Disease.Severity ~ ., data = disease)
```

```{r}
summary(fit_full_add)
```


```{r}
fitted_vs_resid(fit_full_add)
qqplot(fit_full_add)
```

```{r}
n = nrow(disease)
p = length(coef(fit_full_add))
```

```{r}
cook_full_add = cooks.distance(fit_full_add)
outlier_full_add = which(cook_full_add > 4 / n)
cook_full_add[outlier_full_add]
```

```{r}
fit_full_add = lm(Disease.Severity ~ ., data = disease[-outlier_full_add,])
```

```{r}
fitted_vs_resid(fit_full_add)
qqplot(fit_full_add)
```
```{r}
fit_aic_add = step(fit_full_add, direction = "both", trace = 0)
``` 

```{r}
fitted_vs_resid(fit_aic_add)
qqplot(fit_aic_add)
```

```{r}
summary(fit_aic_add)
```


**Log model**

```{r}
fit_full_log_add = lm(log(Disease.Severity) ~ ., data = disease)
```

```{r}
summary(fit_full_log_add)
```

```{r}
fit_bic_log_add = step(fit_full_log_add, direction = "backward", k = log(n), trace = 0)
```

```{r}
summary(fit_bic_log_add)
```

```{r}
fitted_vs_resid(fit_bic_log_add)
qqplot(fit_bic_log_add)
```

```{r}
cook_bic_log = cooks.distance(fit_bic_log_add)
outlier_log_add = which(cook_bic_log > 4 / n)
cook_bic_log[outlier_full_add]
```

```{r}
fit_full_log_add = lm(log(Disease.Severity) ~ ., data = disease[-outlier_log_add, ])
```

```{r}
fit_bic_log_add = step(fit_full_log_add, direction = "backward", trace = 0, k = log(n))
```

```{r}
summary(fit_bic_log_add)
```
```{r}
fitted_vs_resid(fit_bic_log_add)
qqplot(fit_bic_log_add)
```

```{r}
bptest(fit_bic_log_add)
```

```{r}
shapiro.test(resid(fit_bic_log_add))
```

**Log Interaction**

```{r, eval = FALSE}
fit_full_log_int = lm(log(Disease.Severity) ~ . ^ 2, data = disease[-outlier_log_add, ])
```

```{r, eval = FALSE}
fit_bic_log_int = step(fit_full_log_int, direction = "backward", k = log(n), trace = 0)
```

```{r, eval = FALSE}
summary(fit_bic_log_int)
```

```{r, eval = FALSE}
fitted_vs_resid(fit_bic_log_int)
qqplot(fit_bic_log_int)
```

```{r, eval = FALSE}
bptest(fit_bic_log_int)
shapiro.test(resid(fit_bic_log_int))
```

```{r, eval = FALSE}
anova(fit_bic_log_add, fit_bic_log_int)
```

**Boxcox model**



```{r}
boxcox(fit_full_add, plotit = TRUE, lambda = seq(-0.5, 0, by = 0.05))
```

```{r}
fit_boxcox = lm(((Disease.Severity^(-0.25) - 1) / (-0.25)) ~ ., data = disease)
```

```{r}
fit_bic_boxcox = step(fit_boxcox, direction = "backward", trace = 0, k = log(n))
```

```{r}
summary(fit_bic_boxcox)
```

```{r}
cook_boxcox = cooks.distance(fit_bic_boxcox)
outlier_boxcox = which(cook_boxcox > 4 / n)
cook_boxcox[outlier_boxcox]
```
```{r}
fit_boxcox = lm(((Disease.Severity^(-0.25) - 1) / (-0.25)) ~ ., data = disease[-outlier_boxcox,])
```

```{r}
fit_bic_boxcox = step(fit_boxcox, direction = "backward", trace = 0, k = log(n))
```


```{r}
summary(fit_bic_boxcox)
```

```{r}
fitted_vs_resid(fit_bic_boxcox)
qqplot(fit_bic_boxcox)
```

```{r}
bptest(fit_bic_boxcox)
shapiro.test(resid(fit_bic_boxcox))
```


```{r}

all_boxcox_models = summary(regsubsets(((Disease.Severity^(-0.25) - 1) / (-0.25)) ~ ., data = disease))
all_boxcox_models$which
```

```{r}
fit_boxcox_small = lm(((Disease.Severity^(-0.25) - 1) / (-0.25)) ~ PCDH12 + DLG5 + SHISA5 + AF161342 + F2R + PLEKHM1 + PPAN + LOC440104, data = disease[-outlier_log_add,])
```

```{r, eval = FALSE}
anova(fit_boxcox_small, fit_boxcox)
```

